# Data

## Description
1. NYC Dog Licensing Dataset via NYC Open Data

Link: https://data.cityofnewyork.us/Health/NYC-Dog-Licensing-Dataset/nu7n-tubp/about_data

Description: Our primary data source will be the NYC Dog Licensing Dataset from NYC Open Data. This data is provided by the Department of Mental Health and Hygiene.  The data is collected as part of the DOHMH Dog Licensing System, where all NYC dog owners are required to apply for dog licenses.  The data is updated annually and was last updated on February 6th, 2024. Each record represents a unique dog license that was active during the year and contains columns for the animal name, gender, birth year, and breed. We plan to import the data via the downloadable CSV on the NYC Open Data Website. 

The data covers licensing data from 2016-19 and 2023-23; it is missing 2020 and 2021, likely due to the COVID pandemic. While this is obviously a shortcoming, we will still be using the data to visualize overall New York City dog trends, as well as trends before and after the pandemic. We will be sure to not draw any inappropriate or presumptive conclusions given that there is a distinct gap in our data.

The variables in this dataset require basic cleaning. For the dog name and dog breed variables, we code responses like "Not Provided" and "Unknown" as NA. Also, for some visualizations, we may want to define broader categories for the breeds of licensed dogs. For example, we may choose to group these two, "American Pit Bull Terrier/Pit Bull" and "American Pit Bull Mix/ Pit Bull Mix" into one breed "American Pit Bull".

2. Spatial Datasets

Link: (1) NYC ZipCodes at https://data.cityofnewyork.us/Health/Modified-Zip-Code-Tabulation-Areas-MODZCTA-/pri4-ifjk/data 
(2) NYC Open Spaces at https://data.cityofnewyork.us/City-Government/NYC-Planimetric-Database-Open-Space-Parks-/g84h-jbjm

Descriptions: As part of our analysis, we want to visualize the impact of geography on dog ownership in New York City. To do this, we will leverage shapefiles made available by NYC Open Data. The first maps NYC zip codes to MultiPolygons. We can join this with our dog licensing data set which contains a column for zip code. The second spatial dataset maps NYC open spaces to Polygons. Open spaces include malls, parks, golf courses, and others. We will limit this dataset to just open spaces which can be utilized with dogs - namely these land use designations: "Flagship Park", "Community Park", "Neighborhood Park", and "Nature Area".

3. Dog Bite Dataset

Link: https://data.cityofnewyork.us/Health/DOHMH-Dog-Bite-Data/rsgh-akpg/about_data

Description: One of the questions we posed in our introduction was whether dog bite rates have any effect on rates of dog ownership. To answer this question, we will use the Dog Bite Dataset from NYC Open Data. This data was collected by the Department of Health and Mental Hygiene. It is updated as needed and was last updated on February 21, 2024. The NYC Health Code requires all dog bites be reported within 24 hours. Data is received via website, mail, fax or by phone to 311. Each record represents a unique dog bite incident and contains information on dog breed, age, gender and spay/neuter status where available. We plan to import the data via the downloadable CSV on the NYC Open Data Website. 

Our analysis will primarily use this data source to construct counts of bites for different breeds of dogs. Unfortunately, it does not cover all years of our data, as the latest update was for 2022. We will use the range of years that matches to our data for analysis.

The data requires basic cleaning. For the dog breed and other characteristic variables, we code responses like "Not Provided" and "Unknown" as NA. We did discover that this dataset contains a high number of missing values for some of the dog characteristics. We will make note to avoid using these specific characteristics in our analysis.

4. US Census Bureau Data

Link: https://data.census.gov/table?g=040XX00US36_050XX00US36005,36047&y=2023&d=ACS%201-Year%20Estimates%20Selected%20Population%20Profiles

Description: As mentioned in our introduction, we are looking to consider macro social factors and how they may relate to dog ownership. To inform this, we will leverage the latest US Census Bureau Data from 2023. We filter on the counties we are interested in - Staten Island, Kings (Brooklyn), Queens, New York, Bronx. The Census Bureau collects data through variety of sources, including from respondents directly. It also uses existing data from government agencies and commercial entities. The data is released annually. Each column represents one of our selected counties and a demographic within the county (i.e. "Bronx Females"). There are seperate columns for value estimates, percent estimates, and margin of error. Each record represents a different characteristic of that demographic - for example "Median Household Income". Using the Census Bureau website, we can filter the data and then export to CSV.

The dataset also includes a margin of error. Since census data is collected through a sampling method, the dataset provides us with an uncertainty level around the final estimate. Where necessary, we will include this margin of error in our final visualizations. 

## Missing value analysis

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)
library(DataExplorer)

dogs <- read_csv("data/NYC_Dog_Licensing_Dataset_20241119.csv")

dogs$BreedName[dogs$BreedName == "Unknown"] <- NA
dogs$BreedName[dogs$BreedName == "Not Provided"] <- NA

dogs$AnimalName[dogs$AnimalName == "UNKNOWN"] <- NA
dogs$AnimalName[dogs$AnimalName == "NOT PROVIDED"] <- NA
dogs$AnimalName[dogs$AnimalName == "NAME NOT PROVIDED"] <- NA
dogs$AnimalName[dogs$AnimalName == "NA"] <- NA

plot_missing(dogs, title = "Dog License Data")

```
Within the dog licensing data, we observed that around 9% of records are missing values for dog breed. This is within the acceptable range determined by the library and is understandable given that not all dog owners are confident in knowing the breed of their dog (for example, rescue dogs).


```{r, message=FALSE, warning=FALSE}

dog_bite <- read_csv("data/DOHMH_Dog_Bite_Data_20241119.csv")

dog_bite$Gender[dog_bite$Gender == "U"] <- NA
dog_bite$Breed[dog_bite$Breed == "Unknown"] <- NA

plot_missing(dog_bite, title = "Dog Bite Data")

```
Within the dog licensing data, we observed that around 9% of records are missing values for dog breed. This falls within an acceptable range for our analysis. Additionally, multiple columns contained missing data including breed, zip code, gender and age. Gender and age were not within the acceptable range of missing values, so we will avoid using these values in our analysis. 

```{r fig.width=10, fig.height=20, message=FALSE, warning=FALSE}

census_median_income_data <- read_csv("data/ACSST1Y2023.S2001-2024-11-20T015529.csv")

plot_missing(census_median_income_data, title = "NYC Borough Median Income Data")

```
The US Census Bureau Median Income Data we pulled is within the acceptable range of missing values. However, the census data itself does not contain Staten Island, Richmond County, as a possible filter when pulling data. So, Staten Island is not considered in the demographic factor analysis, and we draw conclusions pertaining to the other four boroughs. As mentioned, each column represents a county and a demographic within the county. The rows are categorized into different demographic areas (age, household, etc.), and there are subcategories within larger categorizations. There are separate columns for values, percentages and margins of error. For simplicity, we are considering values and percentages without using margins of error in our analysis.
